Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-25 paper/submission.tex
Removed blue highlighting for final .tex version

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect {\color{blue} statistical causality} between  social sentiment {\color{blue} changes} and cryptocurrency {\color{blue} returns}. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear {\color{blue}statistical causality} to be an order of magnitude larger than {\color{blue}the linear case}.
\end{abstract}

@@

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. {\color{blue} This may be considered as a field of price discovery, which represents a large and fundamental area of financial theory.} In the ideal case {with complete predictive information, \color{blue}from} knowing the movement of one price, the movement of a second can be inferred {\color{blue} exactly. In reality information is incomplete and held asymmetrically, and the ability to reliably obtain this and contribute to price formation is generally rewarded by the market}. In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

  The concept of an entirely peer-to-peer digital currency managed via a distributed ledger was described and applied by Nakamoto in 2008 \cite{nakamoto2008bitcoin}, who named the currency `Bitcoin'. The proposal and subsequent implementation captured the attention of technologists, economists, libertarians and futurists, and spawned numerous adaptations utilising the blockchain technology \cite{aste2017blockchain}, which have come to be known as cryptocurrencies. Trading in these cryptocurrencies has become widely available even to less sophisticated retail investors, and volumes have grown significantly as interest in the currencies has widened. The crypto market is characterised by high volatility which seems to reflect changes in the attitudes of investors. The usage of cryptocurrencies in the traditional economy remains limited, and it is reasonable to assume that prices are in part driven by speculative dynamics, separate to any utility as a medium of exchange or to any revenue-generating process. %Therefore one should expect to observe a larger interplay between sentiment and prices in cryptocurrencies, compared to more fundamental-driven markets such as equities. 
  {\color{blue} One factor which contributes to these dynamics is market sentiment, the estimation of which using social media data is an active area of research}.  We discuss existing literature that identifies signals between social sentiment {\color{blue} data} and equity prices, and so hypothesise that investor sentiment on future cryptocurrency prices may be expected to feed into the short-term price movements via speculation, which may be detected {\color{blue} through the usage of} information-theoretic techniques. This paper tests this hypothesis.

  The relationship between social media sentiment and price {\color{blue} changes} has been explored in the literature for traditional markets and, more recently, for crypto markets as well, including by an author of this paper. For instance, Bollen et al.  \cite{bollen2011twitter} showed that the mood of Twitter messages can be used as a proxy for market sentiment, and that this can show a linear relationship with price movements in US equities. Zheludev \& Aste \cite{zheludev2014can} also performed sentiment analysis using Natural Language Processing (NLP) on Twitter data, to show sentiment is significantly coupled with price movements for a number of instruments issued by S\&P500 firms. Souza \& Aste  used Twitter messages to model market sentiment, and showed the non-linear predictive relationship may be greater than the linear one \cite{souza2016nonlinear}.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price {\color{blue} returns depend} on prior {\color{blue} returns and changes in market sentiment}, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the {\color{blue} change in} social sentiment at $t-k$ on the {\color{blue} returns} at time $t$.

@@

  Though developed relatively recently, information-theoretic methods have been used with success in research across disciplines, to detect information transfer where interventionist approaches are not possible. For example, in neuroscience, Vicente et al. \cite{vicente2011transfer} found transfer entropy to be a superior measure in {\color{blue} detecting Granger causality} in electrophysiological communication than {\color{blue}the auto-regressive formulation}. In climatology, Liang derived from first principles a linear information flow measure, and used this to show that El Ni\~{n}o tends to stabilise the Indian Ocean Dipole \cite{san2014unraveling}. This analysis also detected a causal effect in the other direction; the Indian Ocean Dipole was shown to amplify El Ni\~{n}o oscillations. The technique was used with further success by Stips et al. \cite{stips2016causal} to confirm that recent CO$_2$ emissions show a one-way causality towards global mean temperature anomalies, but that on paleoclimate timescales, this direction is reversed and temperatures drive CO$_2$ levels. Finally, in finance, information transfer was measured between equities indices by Kwon \& Yang, showing that the information transfer was greatest from the US, and greatest towards the APAC region \cite{kwon2008information}. In particular, the S\&P500 was shown to be the strongest driver of other stock indices. In an earlier and somewhat related work, Marschinski \& Kantz   \cite{Marschinski2002} defined and used effective transfer entropy to quantify contagion in financial markets. Similarly, Tungsong et al. \cite{tungsong2018} used transfer entropy to develop upon the previous work by Diebold \& Yilmaz \cite{diebold2009measuring} in quantifying spillover effects between financial markets, generalising the methodology and estimating the time evolution of interconnectedness between financial systems.

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on {\color{blue}Granger-Geweke causality} ({\color{blue}the} linear causality measure) and transfer entropy ({\color{blue}the} non-linear causality measure). In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear {\color{blue}Granger} causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect {\color{blue}statistical} causality in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  \subsection{Linear {\color{blue}Granger} Causality}

@@

  \subsection{Non-Linear {\color{blue}Granger} Causality}

  To detect non-linear {\color{blue}Granger} causality, we apply an information-theoretic approach. Equation \ref{eq:geweke_measure} measures the extent to which the additional information in the lagged variable reduces the variance in the model residuals. Transfer entropy extends this concept by considering the uncertainty, instead of the variance. Adopting Shannon's measure of information \cite{shannon1948}, we can express the uncertainty associated with the continuous random variable $X$ by:

@@

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence{\color{blue}, in expectation, the time-dependence required for there to be any statistical causality}.

  We adopt a shuffling approach producing 100 null-hypothesis transfer entropy values from independently shuffled time-series over the same domain, containing no {\color{blue} causal relationships}. By calculating the mean and standard deviation of the shuffled transfer entropy figures, we estimate the significance of a causal result as the distance between the result and the average shuffled result, standardising by the shuffled standard deviation:

@@

  In order to validate the autoregressive and information-theoretic approaches to detecting {\color{blue}statistical} causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table {\color{blue}2} in the appendix.

@@

  Figure \ref{fig:CLM_confirmation} shows the mean transfer entropy results for 15,000 synthetic data points. We observe that, for this system, the linear method is incapable of detecting the causal relationship; it finds some significant information transfer (at a near-zero level, which we interpret as the linear component), but fails to represent the expected exposure–response relationship and does not correctly identify the expected anticipatory signal in the reverse direction. The information-theoretic method, by contrast, produces results which better represent the increasing coupling strength relationship, and direction of {\color{blue}Granger} causality in the system. The technique also detects {\color{blue}Granger} causality from $Y$ to $X$, for large values of $\alpha$. This effect was also observed by Liang and Hahs \& Pethel, and corresponds to the anticipatory nature of the coupling. Specifically the coupling function $g(x)$ involves repeated application of the update function $f(x)$ from equation \ref{eq:f(x)}, from which it can be seen that $f(f(X_t))$ is equivalent to $X_{t+2}$ so, for large $\alpha$, $Y_t$ will contain increasing amounts of the future information of $X_t$. In fact, at large coupling strengths approaching $\alpha=1$, the observed transfer entropy from $X$ to $Y$ begins to decrease, as more information exists in $Y$ about its future evolution.

@@

  \begin{figure*}[!htb]
    \includegraphics[width=\linewidth]{images/confirming_gbm.pdf}
    \caption{Demonstration that both linear and non-linear transfer entropy methods detect {\color{blue}statistical} causality for linearly coupled synthetic data. The plots are calculated from the mean over 10 realisations using 15,000 data points of the synthetic random walk process from equations \ref{eq:GBM} and \ref{eq:coupled_random_walk} with $\mathbb{E} [\mu]=0.1$ and $\mathbb{E} [\mu^{\prime}]=0.1$, each with standard deviations of $0.1$, and $\mathbb{E} [\sigma]=0.025$ and $\mathbb{E} [\sigma^{\prime}]=0.025$, each with standard deviations of $0.01$.
    Non-linear transfer entropy is calculated using a quantile histogram of 6 classes per dimension. The Z-score of each result is also plotted for both methods. We observe no significant transfer entropy in the non-causal direction $Y \rightarrow X$. For the direction $X \rightarrow Y$, the Z-score of the linear technique is greater by orders of magnitude than the information-theoretic technique. However, both techniques correctly detect the linear transfer entropy, presenting largely equivalent results.}
    \label{fig:GBM_confirmation}

    \includegraphics[width=\linewidth]{images/confirming_logmap.pdf}
    \caption{Demonstration that the non-linear causal relationship in synthetic data generated from equations \ref{eq:coupled_map} and \ref{eq:g(x)} is {\color{blue}correctly} detected only by the non-linear method. The plots are calculated from the mean over 10 realisations over 15,000 data points of the synthetic coupled logistic map process, with $\epsilon = 0.4$. Non-linear transfer entropy is calculated using a histogram of 6 classes per dimension, partitioned using marginal equiquantisation. The Z-score for each result is also plotted for both methods. In the direction $X \rightarrow Y$, {\color{blue} the classical approach is unable to detect the known dependency of Y on X, whilst detecting spurious small signals. By contrast the information-theoretic approach correctly identifies the relationship}.  We note that for $\alpha>0.2$ the information-theoretic method detects information transfer in the other direction. This is observed to increase as $\alpha$ approaches $1$, and corresponds to an expected non-linear anticipatory signal.
    }
    \label{fig:CLM_confirmation}
    %\vspace{-35pt} % Add this to rightsize the float and let it fit on a page
  \end{figure*}
  %\clearpage

@@

%\clearpage
\section{Results with Real Data} \label{s.results}
  Having confirmed that the information-theoretic approach is able to detect both linear and non-linear causalities, we apply the technique to investigate the effect of {\color{blue} changes in} social media sentiment on cryptocurrency {\color{blue} returns, and vice versa}. We also apply the linear method, and compare estimates from both techniques to establish whether linear or non-linear dynamics dominate the observed causal relationship.

@@

  The choice of timescale in aggregating raw sentiment data involves a trade-off: with too fine a timescale, there are not enough messages to estimate sentiment, but a long timescale represents a low-resolution sample which loses too much information about the underlying time series. Since cryptocurrencies are traded in real time on electronic exchanges, we hypothesise that causal signals between sentiment and {\color{blue} returns} operate at sub-hourly timescales; hourly aggregation is the smallest time period available in the data, and so this aggregation of sentiment is used.

@@

  For BTC, in Fig. \ref{fig:BTC_TE}, we detect a strong causative signal, of roughly similar scale in both directions of sentiment to BTC {\color{blue} returns} and in the reverse direction. Analysis of the Z-scores confirms the significance in both directions, with the net information transfer greater in the direction of {\color{blue} returns} to sentiment, with mean net transfer entropy of $-0.0016$ bits.

  LTC, in Fig. \ref{fig:LTC_TE}, shows a similar pattern to BTC, although in this case the predominant direction of information transfer is reversed, with mean net transfer entropy of $0.0039$ bits. The Z-scores show the significance of sentiment to {\color{blue} returns} is consistently greater than in the reverse direction, and also stronger than the other currencies. In addition, there is notably a large linear component to the information transfer, which is unique to LTC.

  XRP, in Fig. \ref{fig:XRP_TE}, shows a clear non-linear causality from sentiment to {\color{blue} returns}, with a significant but smaller causal relationship also in the opposite direction. The mean net transfer entropy is measured for this period as $0.0022$ bits. The signal is more significant from sentiment to {\color{blue} returns}, and especially in the periods ending in 2018.

  ETH, in Fig. \ref{fig:ETH_TE}, shows an interesting and unique behaviour. In particular, there appears to be, initially, a significant signal which collapses in the windows ending around January 2018. The information transfer at its initial peak is greater from {\color{blue} returns} to sentiment, however over the whole period the effect is more bidirectional, with mean net transfer entropy of zero. This suggests a slight phase change, with sentiment driving {\color{blue} returns} after the collapse of the signal, although no net significance is observed after this point in either direction. This strongly indicates another driving mechanism, the effect of which first becomes present around January 2016 (due to 24-month windows). This effect is likely to be associated with the rapid price movements at the time.  %\cite{Ethereum}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC.pdf}
    \caption{Evidence that BTC sentiment and {\color{blue} returns} are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of {\color{blue} returns} to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC.pdf} 
    \caption{Evidence that LTC {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on {\color{blue} returns} than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to {\color{blue} returns}. Peak significance of $Z=42.8$ is observed from sentiment to {\color{blue} returns} in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP.pdf}
    \caption{Evidence that XRP {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to {\color{blue} returns} in the first period, and from {\color{blue} returns} to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to {\color{blue} returns} in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH.pdf} 
    \caption{Evidence that ETH {\color{blue} returns} and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to {\color{blue} returns}, and $Z=19.9$ in the direction of {\color{blue} returns} to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

Information-theoretic and autoregressive techniques were developed and validated on coupled random walks and chaotic logistic maps, confirming the ability of both techniques to detect linear information transfer, and of the information-theoretic technique to detect non-linear information transfer. Following validation, the techniques were applied to historical data describing social media sentiment and cryptocurrency prices to detect information transfer between {\color{blue} changes in} sentiment and price {\color{blue} returns}.

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
    Information transfer between time series is calculated using the asymmetric information-theoretic measure known as transfer entropy. Geweke's autoregressive formulation of Granger causality is used to compute linear transfer entropy, and Schreiber's general, non-parametric, information-theoretic formulation is used to quantify non-linear transfer entropy. 
    We first validate these measures against synthetic data. Then we apply these measures to detect statistical causality between  social sentiment changes and cryptocurrency returns. We validate results by performing permutation tests by shuffling the time series, and calculate the Z-score. We also investigate different approaches for partitioning in nonparametric density estimation which can improve the significance. 
    Using these techniques on sentiment and price data over a 48-month period to August 2018, for four major cryptocurrencies, namely bitcoin (BTC), ripple (XRP), litecoin (LTC) and ethereum (ETH), we detect significant information transfer, on hourly timescales, with greater net information transfer from sentiment to price for XRP and LTC, and instead from price to sentiment for BTC and ETH. We report the scale of non-linear statistical causality to be an order of magnitude larger than the linear case.
\end{abstract}

@@

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. This may be considered as a field of price discovery, which represents a large and fundamental area of financial theory. In the ideal case {with complete predictive information, \color{blue}from} knowing the movement of one price, the movement of a second can be inferred exactly. In reality information is incomplete and held asymmetrically, and the ability to reliably obtain this and contribute to price formation is generally rewarded by the market. In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency price returns.

  The concept of an entirely peer-to-peer digital currency managed via a distributed ledger was described and applied by Nakamoto in 2008 \cite{nakamoto2008bitcoin}, who named the currency `Bitcoin'. The proposal and subsequent implementation captured the attention of technologists, economists, libertarians and futurists, and spawned numerous adaptations utilising the blockchain technology \cite{aste2017blockchain}, which have come to be known as cryptocurrencies. Trading in these cryptocurrencies has become widely available even to less sophisticated retail investors, and volumes have grown significantly as interest in the currencies has widened. The crypto market is characterised by high volatility which seems to reflect changes in the attitudes of investors. The usage of cryptocurrencies in the traditional economy remains limited, and it is reasonable to assume that prices are in part driven by speculative dynamics, separate to any utility as a medium of exchange or to any revenue-generating process. %Therefore one should expect to observe a larger interplay between sentiment and prices in cryptocurrencies, compared to more fundamental-driven markets such as equities. 
  One factor which contributes to these dynamics is market sentiment, the estimation of which using social media data is an active area of research.  We discuss existing literature that identifies signals between social sentiment data and equity prices, and so hypothesise that investor sentiment on future cryptocurrency prices may be expected to feed into the short-term price movements via speculation, which may be detected through the usage of information-theoretic techniques. This paper tests this hypothesis.

  The relationship between social media sentiment and price changes has been explored in the literature for traditional markets and, more recently, for crypto markets as well, including by an author of this paper. For instance, Bollen et al.  \cite{bollen2011twitter} showed that the mood of Twitter messages can be used as a proxy for market sentiment, and that this can show a linear relationship with price movements in US equities. Zheludev \& Aste \cite{zheludev2014can} also performed sentiment analysis using Natural Language Processing (NLP) on Twitter data, to show sentiment is significantly coupled with price movements for a number of instruments issued by S\&P500 firms. Souza \& Aste  used Twitter messages to model market sentiment, and showed the non-linear predictive relationship may be greater than the linear one \cite{souza2016nonlinear}.

@@

  Within a linear causality framework, hypothesising that cryptocurrency price returns depend on prior returns and changes in market sentiment, a Granger causality test can detect the impact of past values of a variable $X_t$ on future values of another variable $Y_t$ \cite{granger1969investigating}. This can be calculated using a vector auto-regressive (VAR) model, which describes the extent to which including past values of $X$ reduces the sum of squared residuals in the regression of $X$ against $Y$, hence estimating the predictive effect, at each lag $k$, of the change in social sentiment at $t-k$ on the returns at time $t$.

@@

  Though developed relatively recently, information-theoretic methods have been used with success in research across disciplines, to detect information transfer where interventionist approaches are not possible. For example, in neuroscience, Vicente et al. \cite{vicente2011transfer} found transfer entropy to be a superior measure in detecting Granger causality in electrophysiological communication than the auto-regressive formulation. In climatology, Liang derived from first principles a linear information flow measure, and used this to show that El Ni\~{n}o tends to stabilise the Indian Ocean Dipole \cite{san2014unraveling}. This analysis also detected a causal effect in the other direction; the Indian Ocean Dipole was shown to amplify El Ni\~{n}o oscillations. The technique was used with further success by Stips et al. \cite{stips2016causal} to confirm that recent CO$_2$ emissions show a one-way causality towards global mean temperature anomalies, but that on paleoclimate timescales, this direction is reversed and temperatures drive CO$_2$ levels. Finally, in finance, information transfer was measured between equities indices by Kwon \& Yang, showing that the information transfer was greatest from the US, and greatest towards the APAC region \cite{kwon2008information}. In particular, the S\&P500 was shown to be the strongest driver of other stock indices. In an earlier and somewhat related work, Marschinski \& Kantz   \cite{Marschinski2002} defined and used effective transfer entropy to quantify contagion in financial markets. Similarly, Tungsong et al. \cite{tungsong2018} used transfer entropy to develop upon the previous work by Diebold \& Yilmaz \cite{diebold2009measuring} in quantifying spillover effects between financial markets, generalising the methodology and estimating the time evolution of interconnectedness between financial systems.

 The rest of the paper is organised as follows. In Section \ref{s.background} we provide a brief background on Granger-Geweke causality (the linear causality measure) and transfer entropy (the non-linear causality measure). In Section \ref{s.method} we describe details of the methodology adopted to quantify and validate linear and non-linear Granger causality, and the techniques used to generate synthetic series of linear and non-linear causal coupling. Section \ref{s.validation} demonstrates that the methodologies correctly detect statistical causality in the linear and non linear case when testing against synthetic data. Results for real data, concerning causality between cryptocurrency price and sentiment, are presented in Section \ref{s.results}. Section \ref{s.conclusions} reports conclusions and perspectives.

@@

  \subsection{Linear Granger Causality}

@@

  \subsection{Non-Linear Granger Causality}

  To detect non-linear Granger causality, we apply an information-theoretic approach. Equation \ref{eq:geweke_measure} measures the extent to which the additional information in the lagged variable reduces the variance in the model residuals. Transfer entropy extends this concept by considering the uncertainty, instead of the variance. Adopting Shannon's measure of information \cite{shannon1948}, we can express the uncertainty associated with the continuous random variable $X$ by:

@@

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence, in expectation, the time-dependence required for there to be any statistical causality.

  We adopt a shuffling approach producing 100 null-hypothesis transfer entropy values from independently shuffled time-series over the same domain, containing no causal relationships. By calculating the mean and standard deviation of the shuffled transfer entropy figures, we estimate the significance of a causal result as the distance between the result and the average shuffled result, standardising by the shuffled standard deviation:

@@

  In order to validate the autoregressive and information-theoretic approaches to detecting statistical causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table 2 in the appendix.

@@

  Figure \ref{fig:CLM_confirmation} shows the mean transfer entropy results for 15,000 synthetic data points. We observe that, for this system, the linear method is incapable of detecting the causal relationship; it finds some significant information transfer (at a near-zero level, which we interpret as the linear component), but fails to represent the expected exposure–response relationship and does not correctly identify the expected anticipatory signal in the reverse direction. The information-theoretic method, by contrast, produces results which better represent the increasing coupling strength relationship, and direction of Granger causality in the system. The technique also detects Granger causality from $Y$ to $X$, for large values of $\alpha$. This effect was also observed by Liang and Hahs \& Pethel, and corresponds to the anticipatory nature of the coupling. Specifically the coupling function $g(x)$ involves repeated application of the update function $f(x)$ from equation \ref{eq:f(x)}, from which it can be seen that $f(f(X_t))$ is equivalent to $X_{t+2}$ so, for large $\alpha$, $Y_t$ will contain increasing amounts of the future information of $X_t$. In fact, at large coupling strengths approaching $\alpha=1$, the observed transfer entropy from $X$ to $Y$ begins to decrease, as more information exists in $Y$ about its future evolution.

@@

  \begin{figure*}[!htb]
    \includegraphics[width=\linewidth]{images/confirming_gbm.pdf}
    \caption{Demonstration that both linear and non-linear transfer entropy methods detect statistical causality for linearly coupled synthetic data. The plots are calculated from the mean over 10 realisations using 15,000 data points of the synthetic random walk process from equations \ref{eq:GBM} and \ref{eq:coupled_random_walk} with $\mathbb{E} [\mu]=0.1$ and $\mathbb{E} [\mu^{\prime}]=0.1$, each with standard deviations of $0.1$, and $\mathbb{E} [\sigma]=0.025$ and $\mathbb{E} [\sigma^{\prime}]=0.025$, each with standard deviations of $0.01$.
    Non-linear transfer entropy is calculated using a quantile histogram of 6 classes per dimension. The Z-score of each result is also plotted for both methods. We observe no significant transfer entropy in the non-causal direction $Y \rightarrow X$. For the direction $X \rightarrow Y$, the Z-score of the linear technique is greater by orders of magnitude than the information-theoretic technique. However, both techniques correctly detect the linear transfer entropy, presenting largely equivalent results.}
    \label{fig:GBM_confirmation}

    \includegraphics[width=\linewidth]{images/confirming_logmap.pdf}
    \caption{Demonstration that the non-linear causal relationship in synthetic data generated from equations \ref{eq:coupled_map} and \ref{eq:g(x)} is correctly detected only by the non-linear method. The plots are calculated from the mean over 10 realisations over 15,000 data points of the synthetic coupled logistic map process, with $\epsilon = 0.4$. Non-linear transfer entropy is calculated using a histogram of 6 classes per dimension, partitioned using marginal equiquantisation. The Z-score for each result is also plotted for both methods. In the direction $X \rightarrow Y$, the classical approach is unable to detect the known dependency of Y on X, whilst detecting small spurious signals. By contrast the information-theoretic approach correctly identifies the relationship.  We note that for $\alpha>0.2$ the information-theoretic method detects information transfer in the other direction. This is observed to increase as $\alpha$ approaches $1$, and corresponds to an expected non-linear anticipatory signal.
    }
    \label{fig:CLM_confirmation}
    %\vspace{-35pt} % Add this to rightsize the float and let it fit on a page
  \end{figure*}
  %\clearpage

@@

%\clearpage
\section{Results with Real Data} \label{s.results}
  Having confirmed that the information-theoretic approach is able to detect both linear and non-linear causalities, we apply the technique to investigate the effect of changes in social media sentiment on cryptocurrency returns, and vice versa. We also apply the linear method, and compare estimates from both techniques to establish whether linear or non-linear dynamics dominate the observed causal relationship.

@@

  The choice of timescale in aggregating raw sentiment data involves a trade-off: with too fine a timescale, there are not enough messages to estimate sentiment, but a long timescale represents a low-resolution sample which loses too much information about the underlying time series. Since cryptocurrencies are traded in real time on electronic exchanges, we hypothesise that causal signals between sentiment and returns operate at sub-hourly timescales; hourly aggregation is the smallest time period available in the data, and so this aggregation of sentiment is used.

@@

  For BTC, in Fig. \ref{fig:BTC_TE}, we detect a strong causative signal, of roughly similar scale in both directions of sentiment to BTC returns and in the reverse direction. Analysis of the Z-scores confirms the significance in both directions, with the net information transfer greater in the direction of returns to sentiment, with mean net transfer entropy of $-0.0016$ bits.

  LTC, in Fig. \ref{fig:LTC_TE}, shows a similar pattern to BTC, although in this case the predominant direction of information transfer is reversed, with mean net transfer entropy of $0.0039$ bits. The Z-scores show the significance of sentiment to returns is consistently greater than in the reverse direction, and also stronger than the other currencies. In addition, there is notably a large linear component to the information transfer, which is unique to LTC.

  XRP, in Fig. \ref{fig:XRP_TE}, shows a clear non-linear causality from sentiment to returns, with a significant but smaller causal relationship also in the opposite direction. The mean net transfer entropy is measured for this period as $0.0022$ bits. The signal is more significant from sentiment to returns, and especially in the periods ending in 2018.

  ETH, in Fig. \ref{fig:ETH_TE}, shows an interesting and unique behaviour. In particular, there appears to be, initially, a significant signal which collapses in the windows ending around January 2018. The information transfer at its initial peak is greater from returns to sentiment, however over the whole period the effect is more bidirectional, with mean net transfer entropy of zero. This suggests a slight phase change, with sentiment driving returns after the collapse of the signal, although no net significance is observed after this point in either direction. This strongly indicates another driving mechanism, the effect of which first becomes present around January 2016 (due to 24-month windows). This effect is likely to be associated with the rapid price movements at the time.  %\cite{Ethereum}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC.pdf}
    \caption{Evidence that BTC sentiment and returns are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of returns to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC.pdf} 
    \caption{Evidence that LTC returns and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on returns than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to returns. Peak significance of $Z=42.8$ is observed from sentiment to returns in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP.pdf}
    \caption{Evidence that XRP returns and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to returns in the first period, and from returns to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to returns in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH.pdf} 
    \caption{Evidence that ETH returns and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to returns, and $Z=19.9$ in the direction of returns to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

Information-theoretic and autoregressive techniques were developed and validated on coupled random walks and chaotic logistic maps, confirming the ability of both techniques to detect linear information transfer, and of the information-theoretic technique to detect non-linear information transfer. Following validation, the techniques were applied to historical data describing social media sentiment and cryptocurrency prices to detect information transfer between changes in sentiment and price returns.

The information-theoretic investigation detected a significant non-linear causal relationship in BTC, LTC and XRP, over multiple timescales and in both directions of sentiment to returns and returns to sentiment. The effect was strongest and most consistent for LTC and XRP, and in both cases the net information transfer was in the direction of sentiment to returns. ETH and BTC were observed to show significant net information transfer in the direction returns to sentiment, although separate dynamics were observed in ETH which see a collapse in significance.
